---
title: "Lab 4"
author: "Your Name Here"
date: ""
output: html_document
---

##### Remember to change the `author: ` field on this Rmd file to your own name.

### Learning objectives

> In today's Lab you will gain practice with the following concepts from today's class:

>- Using the **`tapply`** and **`aggregate`** commands to produce simple tabular summaries, and interpreting the results
- Using the **`by`** command to produce tables showing correlations between variables
- Constructing scatterplots with legends using base graphics
- Drawing connections between tabular and graphical summaries


### Problems

We'll begin by doing all the same data processing as in lecture.
```{r}
library(MASS)

# Assign more descriptive variable names
colnames(birthwt) <- c("birthwt.below.2500", "mother.age", "mother.weight", 
    "race", "mother.smokes", "previous.prem.labor", "hypertension", "uterine.irr", 
    "physician.visits", "birthwt.grams")

# Assign more descriptive factor levels and convert variables to factors as needed
library(plyr)
birthwt <- transform(birthwt, 
            race = as.factor(mapvalues(race, c(1, 2, 3), 
                              c("white","black", "other"))),
            mother.smokes = as.factor(mapvalues(mother.smokes, 
                              c(0,1), c("no", "yes"))),
            hypertension = as.factor(mapvalues(hypertension, 
                              c(0,1), c("no", "yes"))),
            uterine.irr = as.factor(mapvalues(uterine.irr, 
                              c(0,1), c("no", "yes"))),
            birthwt.below.2500 = as.factor(mapvalues(birthwt.below.2500,
                              c(0,1), c("no", "yes")))
            )
```

#### 1. aggregate() vs tapply()

One of the advantages of aggregate() is that it makes it easier to view summary tables when grouping on more than two factors.  

**(a)** Use the `tapply()` function to calculate mean `birthwt.grams` grouped by race, mother's smoking status, and hypertension.

```{r}
# Edit me
```

**One of the cells in the `tapply` output is equal to `NA`.  Explain why.**

> Replace this text with your solution.

**(b)** Repeat part (a), this time using the `aggregate()` function.

```{r}
# Edit me
```

**Do you see an `NA` result?  Explain.**

> Replace this text with your solution.

<br>

#### 2. by() practice

In today's lecture we used the `by()` function to calculate the correlation between mother's age and birth weight broken down by mother's smoking status.  Here's the command that was used:

```{r}
by(data = birthwt[c("birthwt.grams", "mother.age")], 
   INDICES = birthwt["mother.smokes"], 
   FUN = function(x) {cor(x[,1], x[,2])})
```

**(a)** Use the `by()` function to calculate the correlation between mother's age and birth weight broken down by race.

```{r}
# Edit me
```

**(b)** Does the correlation appear to vary by race? Explain.   

> Replace this text with your solution.

<br>

#### 3. Scatterplot practice

In class we made a pretty nice scatterplot using the following command.

```{r, fig.align='center', eval=FALSE}
# Make sure that the color-blind palette is defined
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

points.base <- c(16, 17) # 16: circle, 17: triangle
cols.base <- cbPalette[2:3] # blue and orange points
point.symbols <- points.base[birthwt$mother.smokes] 
point.cols <- cols.base[birthwt$mother.smokes]
with(birthwt, plot(mother.age, birthwt.grams, 
                   main="Birth Weight by Mother's Age", 
                   xlab="Mother's Age (Years)", 
                   ylab="Birth Weight (Grams)",
                   col=point.cols,
                   pch=point.symbols))
legend("bottomright", c("Nonsmoker","Smoker"), col=cols.base, pch=points.base)
```

**(a)** Try to create a similar plot, but instead of using different colors and symbols based on the value of `mother.smokes`, do so based on the value of `race`.  

```{r, fig.height=6, fig.width=6, fig.align='center'}
# Edit me
```

**(b)** Use the `abline` command to add a horizontal line to the plot at `birthwt.grams = 2500`. 

```{r, fig.height=6, fig.width=6, fig.align='center'}
# Edit me
```

**(c)** What does the 2500g horizontal line represent?    In practical terms, what does it mean when a point lies below this line?  

> Replace this text with your solution.
 
**(d)**  There is one isolated point in the top-right corner of the plot.  Describe the race, age and baby birthweight of this point.  Could this isolated point have affected the correlation calculation in Question 2 in any way?  

> Replace this text with your solution.