---
title: "Lab 3"
author: "Your Name Here"
output: html_document
---

##### Remember to change the `author: ` field on this Rmd file to your own name.

### Learning objectives

> In today's Lab you will gain practice with the following concepts from today's class:

>- Using **loops** to iterate through a data set
- Alternatives to loops such as the **`apply`** and **`rowMeans`** commands
- Using the **`tapply`** command to produce simple tabular summaries, and interpreting the results
- Evaluating expressions with the help of the **`with`** command as a way of simplifying syntax

## Problems

```{r}
library(MASS)
```

### 1. Loop practice

**(a)** Write a function called `calculateRowMeans` that uses a **for loop** to calculate the row means of a matrix `x`.

```{r}
# Edit me
```

**(b)** Try out your function on the random matrix `fake.data` defined below.

```{r}
set.seed(12345) # Set seed of random number generator
fake.data <- matrix(runif(800), nrow=25)
```

**(c)** Use the `apply()` function to calculate the row means of the matrix `fake.data`

```{r}
# Edit me
```

**(d)** Compare this to the output of the `rowMeans()` function to check that your calculation is correct.

```{r}
# Edit me
```

### 2. tapply() practice

**(a)** Use `tapply()` on the Cars93 data set to create a table showing the average `Turn.circle` of cars, broken down by vehicle `Type` and `DriveTrain`

```{r}
# Edit me
```

**(b)** Are there any `NA` values in your table?  If so, explain why `NA` values are appearing.

> Replace this text with your solution.

**(c)** Having a car with a small turn radius makes city driving much easier.  Should city drivers opt for smaller vehicles or larger vehicles?

> Replace this text with your solution.

**(d)** Does the vehicle's `DriveTrain` appear to have an impact on turn radius?  

> Replace this text with your solution.

### 3. with() practice

Below is a code chunk that produces box plots of Horsepower against vehicle Type.  

```{r, fig.align='center', fig.width=7, fig.height=5}
boxplot(Cars93$Horsepower ~ Cars93$Type)
```

**(a)** Use the `with()` function to produce the same plot with different syntax.

```{r, fig.align='center', fig.width=7, fig.height=5}
# Edit me
```

**(b)**  Which types of cars tend to be the most powerful (have the highest horsepower)?

> Replace this text with your solution.

**(c)** Which types of cars offer the widest range of power?

> Replace this text with your solution.